Computational methods in landscape ecology

### 2024-09-19 Manuscript.qmd
Remove trailing whitespaces
Updates to general MS structure

keywords: keyword1, keyword2

author:
  - name: Maximilian H.K. Hesselbarth
    orcid: 0000-0003-1125-9918
    email: mhk.hesselbarth@gmail.com
    affil-id: 1
    attributes:
      - corresponding: boolean
  - name: Jakub Nowosad
    orcid: 0000-0002-1057-3721
    email: nowosad.jakub@gmail.com
    affil-id: 2,3
  - name: Martí Bosch
    orcid: 0000-0001-8735-9144
    email: marti.bosch@epfl.ch
    affil-id: 4
  - name: Martin Jung
    orcid: 0000-0002-7569-1390
    email: jung@iiasa.ac.at
    affil-id: 1
  - name: Alida de Flamingh
    orcid: 0000-0003-1223-6654
    email: adeflamingh@gmail.com
    affil-id: 5

affiliations: 
  - id: 1
    name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
  - id: 2
    name: Adam Mickiewicz University, Poznań, Poland
  - id: 3
    name: University of Münster, Münster, Germany
  - id: 4
    name: École polytechnique fédérale de Lausanne, Lausanne, Switzerland
  - id: 5
    name: Carl R. Woese Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA

@@

<!-- MH: Author order is not fixed yet. I just started to list everyone for now. -->

{{< include Sections/01-introduction.qmd >}}
{{< include Sections/02-spatial-patterns.qmd >}}
{{< include Sections/03-connectivity.qmd >}}
{{< include Sections/04-sampling.qmd >}}
{{< include Sections/05-simulations.qmd >}}
<!-- JN: I would suggest switching the order of spatial planning and genetics -->
{{< include Sections/06-spatial-planning.qmd >}}
{{< include Sections/07-genetics.qmd >}}
{{< include Sections/08-discussion.qmd >}}

@@

### References {.unnumbered}

>>>>

keywords: spatial data, spatial patterns, open-source, scientific software

author:
  - name: Maximilian H.K. Hesselbarth
    orcid: 0000-0003-1125-9918
    email: mhk.hesselbarth@gmail.com
    affil-id: 1
  - name: Jakub Nowosad
    orcid: 0000-0002-1057-3721
    email: nowosad.jakub@gmail.com
    affil-id: 2,3
  - name: Gemma Gerber
    orcid: 0000-0002-8722-3113
    email: gerber@iiasa.ac.at
    affil-id: 1
  - name: Alida de Flamingh
    orcid: 0000-0003-1223-6654
    email: adeflamingh@gmail.com
    affil-id: 4
  - name: Martin Jung
    orcid: 0000-0002-7569-1390
    email: jung@iiasa.ac.at
    affil-id: 1
  - name: Craig E. Simpkins
    orcid: 0000-0003-3212-1379
    email: simpkinscraig063@gmail.com
    affil-id: 5
  - name: Martí Bosch
    orcid: 0000-0001-8735-9144
    email: marti.bosch@epfl.ch
    affil-id: 6

affiliations:
  - id: 1
    name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
  - id: 2
    name: Adam Mickiewicz University, Poznań, Poland
  - id: 3
    name: University of Münster, Münster, Germany
  - id: 4
    name: Carl R. Woese Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA
  - id: 5
    name: Auckland Council, Environmental Services Department, Auckland, New Zealand
  - id: 6
    name: École polytechnique fédérale de Lausanne, Lausanne, Switzerland

@@

# Abstract {.unnumbered}

This is a test!

{{< include Sections/01-introduction.qmd >}}
{{< include Sections/02-spatial-patterns.qmd >}}
{{< include Sections/03-connectivity.qmd >}}
<!-- {{< include Sections/04-sampling.qmd >}} -->
{{< include Sections/05-simulations.qmd >}}
{{< include Sections/06-genetics.qmd >}}
{{< include Sections/07-spatial-planning.qmd >}}
{{< include Sections/08-discussion.qmd >}}

# Acknowledgments {.unnumbered}

We are thankful to the anonymous reviewers and the Editors who provided valuable comments to improve the manuscript.

## Funding {.unnumbered}

No external funding was received.

## Author contributions {.unnumbered}

All authors contributed particular sections to the manuscript coordinated by MHKH.
Correspondence to [Maximilian H.K. Hesselbarth](mailto:mhk.hesselbarth@gmail.com)

## Conflict of Interest {.unnumbered}

The authors declare no competing interests.

## Human and Animal Rights and Informed Consent {.unnumbered}

This article does not contain any studies with human or animal subjects performed by any of the authors.

@@

## References {.unnumbered}

### 2024-09-19 Sections/03-connectivity.qmd
Updates to general MS structure

Multiple approaches how to measure connectivity have been developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics], however, quantifying connectivity has proven both technically and conceptually challenging [@Calabrese2004ComparisonshoppersGuide; @Taylor2006LandscapeConnectivity].
Movement data from many individuals is required to infer the elements in the landscape that individuals preferentially move through [@Simpkins2018AssessingPerformance].
While improving technologies have made tracking animal movement at high tempo-spatial resolution available [@Kays2015TerrestrialAnimal; Nathan2022BigdataApproaches], it is still logistically challenging to track high numbers of individuals.
Thus, in practice connectivity is primarily determined by indirect estimations.
@Calabrese2004ComparisonshoppersGuide distinguished three types of estimates: i) structural connectivity determined by the physical attributes of the landscape, ii) potential connectivity as a combination of physical landscape attributes and limited information about dispersal capabilities of species, and iii) actual connectivity related to observations of individuals moving through a landscape.

### Estimating structural connectivity

Landscape metrics are commonly used to quantify various aspects of landscape structure decomposed into composition and configuration.
These metrics describe structural connectivity in terms of total habitat amount, patch size and isolation, or possible connections between habitat patches.
Additionally, more recently there has been an increasing focus on within-patch connectivity using metrics such as the effective mesh size [@Spanowicz2019MeasuringLandscape].
However, landscape metrics are frequently criticized for an ambiguous link to functional connectivity, however, functionality can be partly included by using ecological meaningful metrics, such as the core area [e.g., @Kupfer2012].

### Estimating functional connectivity

#### Resistance surfaces

@@

#### Graph theory

>>>>

Multiple approaches how to measure connectivity have been developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics], however, quantifying connectivity has proven both technically and conceptually challenging [@Calabrese2004ComparisonshoppersGuide; @Taylor2006LandscapeConnectivity].
Movement data from many individuals is required to infer the elements in the landscape that individuals preferentially move through [@Simpkins2018AssessingPerformance].
While improving technologies have made tracking animal movement at high tempo-spatial resolution available [@Kays2015TerrestrialAnimal; @Nathan2022BigdataApproaches], it is still logistically challenging to track high numbers of individuals.
Thus, in practice connectivity is primarily determined by indirect estimations.
@Calabrese2004ComparisonshoppersGuide distinguished three types of estimates: i) structural connectivity determined by the physical attributes of the landscape, ii) potential connectivity as a combination of physical landscape attributes and limited information about dispersal capabilities of species, and iii) actual connectivity related to observations of individuals moving through a landscape.

## Estimating structural connectivity

Landscape metrics are commonly used to quantify various aspects of landscape structure decomposed into composition and configuration.
These metrics describe structural connectivity in terms of total habitat amount, patch size and isolation, or possible connections between habitat patches.
Additionally, more recently there has been an increasing focus on within-patch connectivity using metrics such as the effective mesh size [@Spanowicz2019MeasuringLandscape].
However, landscape metrics are frequently criticized for an ambiguous link to functional connectivity, however, functionality can be partly included by using ecological meaningful metrics, such as the core area [e.g., @Kupfer2012LandscapeEcology].

## Resistance surfaces

@@

## Graph theory

### 2024-09-19 Sections/07-spatial-planning.qmd
Updates to general MS structure

# Spatial planning

Landscape ecology is fun [@Turner1989].
There are many things to explore [@With2019].

>>>>

# Spatial conservation planning {#sec-spatialplanning}

Spatial planning uses decision theory to identify and allocate areas to specific purposes, such as reserve selection through spatial conservation prioritization [@Margules2000SystematicConservation; @Kukkala2013CoreConcepts; @Jung2024AssessmentState].
In the context of landscape-ecology, it provides a pattern-process link<!-- (see section @sec-spatialpattern) -->, and outcomes related to decision making by distinguishing between structure, function and scales [@Hersperger2021LandscapeEcological].
As an integrative approach most spatial planning approaches can benefit from various computational advances in landscape ecology, for example through using simulation model outputs <!-- (see section @sec-simulation) --> as input features  <!-- (see also @daigle2020operationalizing) -->, there have been several recent advances in considering landscape patterns directly in spatial planning.

Generally, spatial patterns influence planning outcomes because of their compactness, contiguity and connectedness.
Furthermore, aggregation can be important because of ecological or practical reasons, for example by ensuring that selected areas satisfy minimum patch size constraints [@Smith2010ApproachEnsuring; @Justeau-Allaire2021ConstrainedOptimization].
Novel computational approaches make use of graph theory to identify valuable landscape areas in terms of compact core habitat and reducing boundary exposure [@Weerasena2023SequentialApproach].
Similar, contiguity and connectivity are generally considered important to ensure that landscapes or habitat patches are connected, such as river networks [@Beger2010IncorporatingAsymmetric].
In particular, recently the direct consideration of connectivity <!-- (see also section @sec-connectivity) --> has received increasing attention in spatial conservation planning and new computational methods are developed [@Hanson2019ConventionalMethods; @Daigle2020OperationalizingEcological; @Hanson2022ComparisonApproaches; @Beger2022DemystifyingEcological].

Particular noteworthy are two advances that incorporate landscape metrics in spatial planning.
One is the use of graph theory in linear programming to guide conservation and restoration efforts [@Justeau-Allaire2021ConstrainedOptimization; @Hamonic2023CumulativeEffects].
These spatial networks approaches have the potential to be provide more cost-effective and precise solutions to design reserve networks [@Hamonic2023CumulativeEffects].
For example, @Justeau-Allaire2021ConstrainedOptimization used landscape ecology theory and metrics to spatially optimize restoration efforts, aiming to maximize broader landscape connectivity in New Caledonia.

A second main development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathway.
Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
Additionally, it can be applied to identify best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation], such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions, as well as the scalability of spatial-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
However, this comes with the drawback of reduced interpretability.

<!-- ## Open-source software in spatial conservation planning -->

Multiple open-source software solutions have been developed to support integration of landscape ecology theory and indicators into spatial planning.
Most software are based on linear or constrained programming approaches requiring a solver to create any outputs.
Unfortunately there is still a gap between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing].

The *prioritizr* R package [@Hanson2024PrioritizrSystematic] supports a range of compactness and contiguity constraints, penalties, and generally allows to flexible create planning problems in multiple configurations.
Similarly, the *coco* R package uses connectivity constraints in an integer linear planning framework to identify potential outcomes that minimize further fragmentation of the landscape [@vanMantgem2023CocoConservation].
The *restoptr* R package uses constrained programming to identify restoration areas that would most effectively contribute to improved landscape connectivity [@Justeau-Allaire2023RestoptrPackage].
Innovatively, it uses landscape ecology indices not as constraints, but as specific targets, e.g., where one could optimize the effective mesh size in a region.
The *CAPTAIN* Python software suite implements reinforcement learning for area-based conservation efforts [@Silvestro2022ImprovingBiodiversity].
However, many spatial planning software still requires substantial prior knowledge on spatial optimization and statistics, which can create a barrier to more widespread adoption.
